A Multiple-Set Overlapping-Domain Decomposed Monte Carlo Synthetic Acceleration Method for Linear Systems

### 2013-09-05 doc/sna2013/paper/slattery_evans_wilson.tex
adding parallel efficiency measures to sna paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}  % allows inclusion of graphics
\usepackage{booktabs}  % nice rules (thick lines) for tables
\usepackage{microtype} % improves typography for PDF
\usepackage{tabls}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[mathcal]{euscript}
\usepackage{tmath}

@@

\abstract{We present a novel multiple-set overlapping-domain
  decomposed strategy for parallelizing the Monte Carlo Synthetic
  Acceleration (MCSA) method. MCSA methods use the Neumann-Ulam class
  of Monte Carlo solvers for linear systems to accelerate the fixed
  point method on which the Neumann-Ulam solvers are based. To
  effectively parallelize MCSA methods requires the parallelization of
  the underlying Neumann-Ulam solvers. To do this in a domain
  decomposed environment, we borrow strategies traditionally
  implemented in Monte Carlo particle transport to parallelize the
  problem. The multiple-set overlapping-domain decomposition algorithm
  is presented along with parallel scaling data for the resulting
  implementation using the Titan Cray XK7 machine at Oak Ridge
  National Laboratory.}

@@

@@

The \textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA)
method is defined as:
\begin{subequations}
  \begin{gather}
    \label{eq:mcsa_1}
    \mathbf{r}^{k} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k}\:,\\
    \label{eq:mcsa_2}
    \mathbf{x}^{k+1/2} = \mathbf{x}^k + \mathbf{r}^k\:,\\
    \label{eq:mcsa_3}
    \mathbf{r}^{k+1/2} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k+1/2}\:,\\
    \label{eq:mcsa_4}
    \mathbf{A}\delta\mathbf{x}^{k+1/2} = \mathbf{r}^{k+1/2}\:,\\
    \label{eq:mcsa_5}
    \mathbf{x}^{k+1} = \mathbf{x}^{k+1/2} + \delta \mathbf{x}^{k+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where $\mathbf{r}$ is the residual vector, $\delta\mathbf{x}$ is the
Monte Carlo correction vector, and $k$ indicates the iteration
index. Here, a Neumann-Ulam Monte Carlo method is used to generate the
solution correction in Eq~\ref{eq:mcsa_4} from the residual and
Richardson's iteration in the first step has been rewritten as a
residual correction. Using Monte Carlo in this way achieves the same
effect as Halton's method, decoupling its convergence rate from the
overall convergence rate of the method. Here, the approximate Monte
Carlo solution is not driven to a particular convergence as it merely
supplies a correction for the initial guess generated by Richardson's
iteration. Rather, only a set number of random walk permuations are
required using the Neumann-Ulam method to generate the correction. In
addition to the Monte Carlo solver parameters dictating the number of
histories and weight cutoff, the outer MCSA iterations also have the
following stopping criteria:
\begin{equation}
  ||\mathbf{r}||_\infty < \epsilon \ ||\mathbf{b}||_\infty\:,
  \label{eq:mcsa_stopping_criteria}
\end{equation}
where $\epsilon$ is a user-defined parameter. We refer the reader to
\cite{evans_monte_2012} and \cite{evans_monte_2009} for a more
detailed explanation and analysis of MCSA and the Neumann-Ulam
methods.

@@

Once the nearest-neighbor communication sequence has been completed
for all histories, the completion of transport sequence is
required. To stop transport, an asynchronous binary communication tree
as presented for the same 9 subdomain example in
Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
process has one parent process (except for the MASTER process 0) to
which it will non-blocking send the number of histories that
terminated their transport procedure by weight cutoff in the local
domain. Equivalently, each process has up to two child processes from
which it will receive their completed number of histories with a
non-blocking receive operation. Setting up a tree in this manner lets
the completed history tally be updated incrementally and funneled to
the root process. Because we are solving fixed source problems with
the Neumann-Ulam algorithm without any type of variance reduction that
may generate more histories that we started with, once the root
process completion sum tallies to the number of histories in the
source, transport is complete. Once this occurs, the root process
non-blocking sends the stop message to its children and each process
non-blocking receives a stop message from its parent as shown in
Figure~\ref{fig:binary_comm_tree}. The stop message is then propagated
up the tree in the same manner.

\begin{figure}[h!]
  \begin{center}
    \scalebox{0.6}{
      \input{binary_comm_tree.pdftex_t} }
  \end{center}
  \caption{Binary communication tree for coordinating the end of a
    parallel Neumann-Ulam solve. \textit{Each child process reports to
      its parents how many histories completed within its domain. When
      the root process sums all complete histories, it forwards the
      stop signal to its children which in turn forward the message up
      the tree. The subdomains are indexed by an integer.}}
  \label{fig:binary_comm_tree}
\end{figure}

@@

\subsection{Parallel MCSA Algorithm}
We utilize parallel matrix-vector multiply, vector update, and vector
norm operations as oultined in \cite{saad_iterative_2003} for a
parallel MCSA implementation with these computations required in
Eqs~(\ref{eq:mcsa_1}), (\ref{eq:mcsa_2}), (\ref{eq:mcsa_4}), and
(\ref{eq:mcsa_5}). In the first step, a parallel vector update is used
to apply the residual to the previous iterate's solution in a
Richardson iteration. In the next step and Eq~(\ref{eq:mcsa_5}), the
residual is computed by a parallel matrix-vector multiply and vector
update. Once the correction is computed with a parallel adjoint
Neumann-Ulam solve, this correction is applied to the solution with a
parallel vector update. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parallel Scaling Studies}
We will next measure performance in various scaling studies using an
implementation of the parallel MCSA algorithm to assess its quality
using a simple two-dimensional neutron diffusion problem in a
homogeneous medium. For the scaling results presented here, all
parallel MCSA results were computed using the Monte Carlo Linear
Solvers\footnote{MCLS is an open source C++ library built on the
  Trilinos\cite{heroux_overview_2005} framework and can be downloaded
  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
specifically for this work. For every scaling study presented, the
calculations were performed 3 times with the average values reported
for wall time and those average values used for the efficiency
computations. For all cases, both strong and weak scaling studies were
performed. For strong scaling, the objective is to then decrease the
runtime of a fixed size global problem as a linear function of
$P$. Based on this, we can define the strong scaling \textit{absolute
  efficiency} as:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership class
machines, it is often the case that a problem size large enough to
effectively execute on $O(10,000)$ cores or larger is significantly
larger than any problem that may be solved by a single core due to
memory restrictions. Therefore, we consider absolute scaling for a
base case where $Q$ processes are used instead of a serial
computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
For weak scaling, For strong scaling, the objective is to then decrease the runtime of a
fixed size global problem as a linear function of $P$. Based on this,
we can define the strong scaling \textit{absolute efficiency} as:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership class
machines, it is often the case that a problem size large enough to
effectively execute on $O(10,000)$ cores or larger is significantly
larger than any problem that may be solved by a single core due to
memory restrictions. Therefore, we consider absolute scaling for a
base case where $Q$ processes are used instead of a serial
computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
All parallel scaling studies presented in this work were performed on
the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
Facility located at Oak Ridge National Laboratory.

@@

\subsection{Overlapping Domain Decomposition}
\label{subsec:overlapping_domain_decomp}

In the next set of scaling studies, we will explore how adding overlap
to the subdomains in the problem affects run times and parallel
performance as compared to the case with pure domain decomposition. To
do this, we utilize the set of analytic relations developed in
Appendix~\ref{chap:parallel_theory} to design these numerical
experiments. For the neutron diffusion problem used in these scaling
studies, the linear operator had a spectral radius of 0.787 and a
weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
length of 19.22 transitions for any given history in the
problem. Using the mean-chord approximation given by
Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
5.3\% of the histories born in a particular domain to leak out after
the first transport stage. By adding overlap, we aim to reduce this
fraction of communicated histories and therefore boost the parallel
performance of the algorithm.

@@

For the performance metrics, we may use the same weak and strong
scaling efficiency metrics that we used for the pure domain
decomposition case. As we are comparing to the case without overlap
and are looking to boost performance relative to this case, for both
weak and strong scaling calculations we will use the 16-core case
without overlap as the reference computation for all efficiency
measures reported. The absolute efficiency results from the strong
scaling exercise are given in
Figure~\ref{fig:titan_strong_overlap}. Aside from the boosting of
efficiencies at 2,048 cores, little evidence of strong scaling
improvement is evident for all levels of overlap tested. To further
explore any potential benefits of adding overlap into the system,
Figure~\ref{fig:titan_strong_overlap_diff} gives the difference in the
efficiencies between all overlap cases and the reference case without
overlap at each core count. At smaller numbers of cores, there is no
observed benefit to using overlap to improve strong scaling. The
improvements noted at 2,048 cores are also present here and account
for the best efficiency improvements for all core counts tested. At
16,384 cores, using 2 levels of overlap did give a 6.4\% boost in
efficiency while at 1,024 cores this same amount of overlap actually
reduced the efficiency. Beyond 2 levels of overlap, efficiency levels
begin to fall with 5 levels of overlap performing the worst for all
cases. From this, we don't expect adding any more overlap will improve
the scalability of the algorithm and the best strong scaling
performance was observed when the number of overlap levels is
approximately equal to the root-mean-squared number of states we
expect a history to move from its starting point as computed by
Eq~(\ref{eq:discrete_distance}).

@@

\subsection{Multiple Set Domain Decomposition}
\label{subsec:ms_decomposition}

We next look to apply replication to the problem in the form of
multiple sets. As the previous analysis showed overlap to be of little
assistance in improving scalability, we only apply multiple sets to
the problem with pure domain decomposition within the sets for these
scaling studies. When adding sets, we must decide how many stochastic
histories to use to generate the correction vector in the MCSA
iteration. For this work we consider two choices; the split case and
the fully replicated case. As an example of these cases, consider a
single set problem where 20,000 histories are used in the Monte Carlo
solve. If we choose the split case, then solving the same problem with
2 sets would give 10,000 histories in each set and 4 sets would give
5,000 histories in each set. Splitting the histories in this way
provides the same global number of histories used to compute the
correction in the single set problem and we should therefore expect
the same iterative performance at a reduced time per iteration and
thus reduced time to solution. For the replicated history case, each
additional set in the problem will use just as many histories as the
single set case. Using 2 sets with fully replicated histories for our
example problem then has each set computing 20,000 histories with
40,000 global histories in the computation. For the 4 set case, the
number of global histories is increased to 80,000. By increasing the
global number of histories, the replicated cases should approximately
maintain time per iteration but should decrease the number of
iterations required to converge and subsequently reduce time to
solution.

\subsubsection{Strong Scaling with Multiple Sets}
\label{subsubsec:ms_strong}

For the strong scaling case, recall that the objective is to decrease
the time to solution for a given global problem size $N$ by increasing
$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
(\ref{eq:algorithmic_efficiency}), and
(\ref{eq:implementation_efficiency}) to compute the various strong
scaling efficiencies, but we must modify the reference computation in
each of these formulas to reflect our objective. For all efficiency
computations, the 16 core single set computation will be used as the
reference case. By adding more cores to the problem, either by
replicating with multiple sets, or decreasing the local problem size,
any computational resources added to the problem should aim to result
in speedup of this reference case. If this speedup for one technique
of adding cores is less than another, then that technique is deemed a
less efficient use of those computational resources.

@@

\subsubsection{Weak Scaling with Multiple Sets}
\label{subsubsec:ms_weak}

Next, we continue to a weak scaling study of the multiple sets
problem. In the same way as the strong scaling study, we will modify
the pure domain decomposition study to account for the additional
replication. For each calculation in the single set study, that
problem will be replicated either 2 or 4 times with both split and
replicated histories to assess the effects on scalability and
run time. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
the multiple sets weak scaling study. Of primary importance here is
the fact that adding sets to the problem with both split and
replicated histories actually decreases the time to solution. This was
not observed in the strong scaling case until much larger core counts
as the multiple set computations hit the strong scaling wall after the
single set computation.

@@

For a multiple set analysis of the weak scaling timing results, we
again have to consider that adding sets to the problem is actually a
form of strong scaling. Consider a single set problem ($S=1$) for a
given $P$ and $N$. When that problem is now solved using 2 sets we
have $S=2$, $N$, and $2P_s$ where $P_s$ is the number of processors in
a single set. For any number of sets we then have a global problem
size of $N$ while the number of processors is $SP_s$. Therefore,
$N/P$ is no longer fixed in the weak scaling study but rather
$N/(SP_s)$ is fixed. As a result, we modify the weak scaling absolute
efficiency relationship to account for this fact:
\begin{equation}
\eta_{weak}(M|N,P|Q,S) = \frac{T(M,Q)}{S \times T(N,P)}\:,
  \label{eq:ms_weak_efficiency}
\end{equation}
where now the absolute efficiency is a function of the number of sets
in the problem and again $M/Q = N/P$. As with the strong scaling
study, we will use the 16-core single set case as the reference
computation for efficiencies. Interestingly,
Eq~(\ref{eq:ms_weak_efficiency}) has a very similar form to the strong
scaling efficiency given by Eq~(\ref{eq:strong_scaling_absolute}),
representative of the strong scaling nature of adding additional
sets. For problems where adding sets improves time to solution, this
new weak scaling formulation accounts for how efficiently those extra
resources are incorporated into the problem.

@@

It was found the MCSA can indeed be effectively parallelized using the
multiple-set overlapping-domain decomposition algorithm borrowed from
the reactor physics community.  The new algorithm was tested in a wide
variety of parallel scaling studies on the Titan Cray XK7 machine at
the Oak Ridge Leadership Computing Facility. To test the algorithm at
high levels of concurrency, up to 65,356 cores were used in strong
scaling exercises and 131,072 cores used in weak scaling exercises
using the neutron diffusion problem.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}  % allows inclusion of graphics
\usepackage{booktabs}  % nice rules (thick lines) for tables
\usepackage{microtype} % improves typography for PDF
\usepackage{tabls}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[mathcal]{euscript}
\usepackage{tmath}
\usepackage[section]{placeins}

@@

\abstract{We present a multiple-set overlapping-domain decomposed
  strategy for parallelizing the Monte Carlo Synthetic Acceleration
  (MCSA) method. MCSA methods use the Neumann-Ulam class of Monte
  Carlo solvers for linear systems to accelerate the fixed point
  method on which the Neumann-Ulam solvers are based. To effectively
  parallelize MCSA methods requires the parallelization of the
  underlying Neumann-Ulam solvers. To do this in a domain decomposed
  environment, we borrow strategies traditionally implemented in Monte
  Carlo particle transport to parallelize the problem. The
  multiple-set overlapping-domain decomposition algorithm is presented
  along with parallel scaling data for the resulting implementation
  using the Titan Cray XK7 machine at Oak Ridge National Laboratory.}

@@

In this paper, we will present the Monte Carlo methods of Neumann and
Ulam and then outline Monte Carlo Synthetic Acceleration
methods. Next, we present a parallel algorithm for MCSA based on the
work of Brunner and Brantley \cite{brunner_efficient_2009} and Wagner
et. al. \cite{wagner_hybrid_2010}. We then perform a series of
parallel scaling studies to characterize the parallel
algorithm. Finally, we summarize the work and offer future work that
may improve the algorithm.

@@

The \textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA)
method is defined as:
\begin{subequations}
  \begin{gather}
    \label{eq:mcsa_1}
    \mathbf{r}^{k} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k}\:,\\
    \label{eq:mcsa_2}
    \mathbf{x}^{k+1/2} = \mathbf{x}^k + \mathbf{r}^k\:,\\
    \label{eq:mcsa_3}
    \mathbf{r}^{k+1/2} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k+1/2}\:,\\
    \label{eq:mcsa_4}
    \mathbf{A}\delta\mathbf{x}^{k+1/2} = \mathbf{r}^{k+1/2}\:,\\
    \label{eq:mcsa_5}
    \mathbf{x}^{k+1} = \mathbf{x}^{k+1/2} + \delta \mathbf{x}^{k+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where $\mathbf{r}$ is the residual vector, $\delta\mathbf{x}$ is the
Monte Carlo correction vector, and $k$ indicates the iteration
index. Here, a Neumann-Ulam Monte Carlo method is used to generate the
solution correction in Eq~\ref{eq:mcsa_4} from the residual and
Richardson's iteration in the first step has been rewritten as a
residual correction. Using Monte Carlo in this way achieves the same
effect as Halton's method, decoupling its convergence rate from the
overall convergence rate of the method. Here, the approximate Monte
Carlo solution is not driven to a particular convergence as it merely
supplies a correction for the initial guess generated by Richardson's
iteration. Rather, only a set number of random walk permutations are
required using the Neumann-Ulam method to generate the correction. In
addition to the Monte Carlo solver parameters dictating the number of
histories and weight cutoff, the outer MCSA iterations also have the
following stopping criteria:
\begin{equation}
  ||\mathbf{r}||_\infty < \epsilon \ ||\mathbf{b}||_\infty\:,
  \label{eq:mcsa_stopping_criteria}
\end{equation}
where $\epsilon$ is a user-defined parameter. We refer the reader to
\cite{evans_monte_2012} and \cite{evans_monte_2009} for a more
detailed explanation and analysis of MCSA and the Neumann-Ulam
methods.

@@

Once the nearest-neighbor communication sequence has been completed
for all histories, the completion of transport sequence is
required. To stop transport, an asynchronous binary communication tree
as presented for the same 9 subdomain example in
Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
process has one parent process (except for the MASTER process 0) to
which it will non-blocking send the number of histories that
terminated their transport procedure by weight cutoff in the local
domain. Equivalently, each process has up to two child processes from
which it will receive their completed number of histories with a
non-blocking receive operation. Setting up a tree in this manner lets
the completed history tally be updated incrementally and funneled to
the root process. Because we are solving fixed source problems with
the Neumann-Ulam algorithm without any type of variance reduction that
may generate more histories that we started with, once the root
process completion sum tallies to the number of histories in the
source, transport is complete. Once this occurs, the root process
non-blocking sends the stop message to its children and each process
non-blocking receives a stop message from its parent as shown in
Figure~\ref{fig:binary_comm_tree}. The stop message is then propagated
up the tree in the same manner.
\begin{figure}[h!]
  \begin{center}
    \scalebox{0.6}{
      \input{binary_comm_tree.pdftex_t} }
  \end{center}
  \caption{Binary communication tree for coordinating the end of a
    parallel Neumann-Ulam solve. \textit{Each child process reports to
      its parents how many histories completed within its domain. When
      the root process sums all complete histories, it forwards the
      stop signal to its children which in turn forward the message up
      the tree. The subdomains are indexed by an integer.}}
  \label{fig:binary_comm_tree}
\end{figure}

@@

To demonstrate the algorithm, consider the example adapted from
Mervin's work with Wagner and others in the same area
\cite{mervin_variance_2012} and presented in
Figure~\ref{fig:msod_example}.
\begin{figure}[h!]
  \begin{center}
    \scalebox{0.8}{ \input{msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as that represented by the path of particle
C. In this case, particle C is scattering between two adjacent
domains, incurring a large latency cost for a single
particle. Finally, with particle B we observe that 0.5 domain overlap
will still not eliminate all communications. However, if 1 domain
overlap were used, the entire geometry shown in
Figure~\ref{fig:msod_example} would be contained on the source
processor and therefore transport of all 3 particles without
communication would occur.

\subsection{Parallel MCSA Algorithm}
We utilize parallel matrix-vector multiply, vector update, and vector
norm operations as outlined in \cite{saad_iterative_2003} for a
parallel MCSA implementation with these computations required in
Eqs~(\ref{eq:mcsa_1}), (\ref{eq:mcsa_2}), (\ref{eq:mcsa_4}), and
(\ref{eq:mcsa_5}). In the first step, a parallel vector update is used
to apply the residual to the previous iterate's solution in a
Richardson iteration. In the next step and Eq~(\ref{eq:mcsa_5}), the
residual is computed by a parallel matrix-vector multiply and vector
update. Once the correction is computed with a parallel adjoint
Neumann-Ulam solve, this correction is applied to the solution with a
parallel vector update. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parallel Scaling Studies}
We will next measure performance in various scaling studies using an
implementation of the parallel MCSA algorithm to assess its quality
using a simple two-dimensional neutron diffusion problem in a
homogeneous medium. For the scaling results presented here, all
parallel MCSA results were computed using the Monte Carlo Linear
Solvers\footnote{MCLS is an open source C++ library built on the
  Trilinos\cite{heroux_overview_2005} framework and can be downloaded
  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
specifically for this work. For every scaling study presented, the
calculations were performed 3 times with the average values reported
for wall time and those average values used for the efficiency
computations.

For all cases, both strong and weak scaling studies were
performed. For strong scaling, the objective is to then decrease the
runtime of a fixed size global problem as a linear function of
$P$. Based on this, we can define the strong scaling \textit{absolute
  efficiency} as \cite{keyes_how_1999}:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership class
machines, it is often the case that a problem size large enough to
effectively execute on $O(10,000)$ cores or larger is significantly
larger than any problem that may be solved by a single core due to
memory restrictions. Therefore, we consider absolute scaling for a
base case where $Q$ processes are used instead of a serial
computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
For weak scaling, the ratio $N/P$ is fixed such that the local problem
size remains the same for all values of $P$. In this case, the
objective is to maintain a constant wall time while both $N$ and $P$
grow as a fixed ratio. In this case, we define the absolute efficiency
as \cite{keyes_how_1999}:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed.

The absolute efficiency measures presented here only considers the
total time to solution. Often, that total time to solution may be
changing for an iterative method due to the fact that the number of
iterations required to converge may be changing as a function of
$P$. This will often happen in cases where modifying $N$ changes the
spectral character of the problem, often making it stiffer and
therefore more difficult to solve when $N$ increases. Therefore we
also consider \textit{algorithmic efficiency} which considers this
potential change in iteration count \cite{keyes_how_1999}:
\begin{equation}
  \eta_{alg}(N,P|Q) = \frac{I(N,Q)}{I(N,P)}\:,
  \label{eq:algorithmic_efficiency}
\end{equation}
where $I(N,P)$ is the number of iterations required to converge a
problem of global size $N$ computed using $P$ processes.  All parallel
scaling studies presented in this work were performed on the Titan
Cray XK7 machine at the Oak Ridge Leadership Computing Facility
located at Oak Ridge National Laboratory.

@@

\subsection{Overlapping Domain Decomposition}
\label{subsec:overlapping_domain_decomp}
In the next set of scaling studies, we will explore how adding overlap
to the subdomains in the problem affects run times and parallel
performance as compared to the case with pure domain decomposition. To
do this, we utilize the set of analytic relations developed in
Appendix~\ref{chap:parallel_theory} to design these numerical
experiments. For the neutron diffusion problem used in these scaling
studies, the linear operator had a spectral radius of 0.787 and a
weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
length of 19.22 transitions for any given history in the
problem. Using the mean-chord approximation given by
Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
5.3\% of the histories born in a particular domain to leak out after
the first transport stage. By adding overlap, we aim to reduce this
fraction of communicated histories and therefore boost the parallel
performance of the algorithm.

@@

For the performance metrics, we may use the same weak and strong
scaling efficiency metrics that we used for the pure domain
decomposition case. As we are comparing to the case without overlap
and are looking to boost performance relative to this case, for both
weak and strong scaling calculations we will use the 16-core case
without overlap as the reference computation for all efficiency
measures reported. The absolute efficiency results from the strong
scaling exercise are given in
Figure~\ref{fig:titan_strong_overlap}. Aside from the boosting of
efficiencies at 2,048 cores, little evidence of strong scaling
improvement is evident for all levels of overlap tested.

To further explore any potential benefits of adding overlap into the
system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
difference in the efficiencies between all overlap cases and the
reference case without overlap at each core count. At smaller numbers
of cores, there is no observed benefit to using overlap to improve
strong scaling. The improvements noted at 2,048 cores are also present
here and account for the best efficiency improvements for all core
counts tested. At 16,384 cores, using 2 levels of overlap did give a
6.4\% boost in efficiency while at 1,024 cores this same amount of
overlap actually reduced the efficiency. Beyond 2 levels of overlap,
efficiency levels begin to fall with 5 levels of overlap performing
the worst for all cases. From this, we don't expect adding any more
overlap will improve the scalability of the algorithm and the best
strong scaling performance was observed when the number of overlap
levels is approximately equal to the root-mean-squared number of
states we expect a history to move from its starting point as computed
by Eq~(\ref{eq:discrete_distance}).

@@

\subsection{Multiple Set Domain Decomposition}
\label{subsec:ms_decomposition}
We next look to apply replication to the problem in the form of
multiple sets. As the previous analysis showed overlap to be of little
assistance in improving scalability, we only apply multiple sets to
the problem with pure domain decomposition within the sets for these
scaling studies. When adding sets, we must decide how many stochastic
histories to use to generate the correction vector in the MCSA
iteration. For this work we consider two choices; the split case and
the fully replicated case. As an example of these cases, consider a
single set problem where 20,000 histories are used in the Monte Carlo
solve. If we choose the split case, then solving the same problem with
2 sets would give 10,000 histories in each set and 4 sets would give
5,000 histories in each set. Splitting the histories in this way
provides the same global number of histories used to compute the
correction in the single set problem and we should therefore expect
the same iterative performance at a reduced time per iteration and
thus reduced time to solution. For the replicated history case, each
additional set in the problem will use just as many histories as the
single set case. Using 2 sets with fully replicated histories for our
example problem then has each set computing 20,000 histories with
40,000 global histories in the computation. For the 4 set case, the
number of global histories is increased to 80,000. By increasing the
global number of histories, the replicated cases should approximately
maintain time per iteration but should decrease the number of
iterations required to converge and subsequently reduce time to
solution.

\subsubsection{Strong Scaling with Multiple Sets}
\label{subsubsec:ms_strong}
For the strong scaling case, recall that the objective is to decrease
the time to solution for a given global problem size $N$ by increasing
$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
(\ref{eq:algorithmic_efficiency}), and
(\ref{eq:implementation_efficiency}) to compute the various strong
scaling efficiencies, but we must modify the reference computation in
each of these formulas to reflect our objective. For all efficiency
computations, the 16 core single set computation will be used as the
reference case. By adding more cores to the problem, either by
replicating with multiple sets, or decreasing the local problem size,
any computational resources added to the problem should aim to result
in speedup of this reference case. If this speedup for one technique
of adding cores is less than another, then that technique is deemed a
less efficient use of those computational resources.

@@

\subsubsection{Weak Scaling with Multiple Sets}
\label{subsubsec:ms_weak}
Next, we continue to a weak scaling study of the multiple sets
problem. In the same way as the strong scaling study, we will modify
the pure domain decomposition study to account for the additional
replication. For each calculation in the single set study, that
problem will be replicated either 2 or 4 times with both split and
replicated histories to assess the effects on scalability and
run time. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
the multiple sets weak scaling study. Of primary importance here is
the fact that adding sets to the problem with both split and
replicated histories actually decreases the time to solution. This was
not observed in the strong scaling case until much larger core counts
as the multiple set computations hit the strong scaling wall after the
single set computation.

@@

For a multiple set analysis of the weak scaling timing results, we
again have to consider that adding sets to the problem is actually a
form of strong scaling. Consider a single set problem ($S=1$) for a
given $P$ and $N$. When that problem is now solved using 2 sets we
have $S=2$, $N$, and $P=2P_s$ where $P_s$ is the number of processors in
a single set. For any number of sets we then have a global problem
size of $N$ while the number of processors is $SP_s$. Therefore,
$N/P$ is no longer fixed in the weak scaling study but rather
$N/(SP_s)$ is fixed. As a result, we modify the weak scaling absolute
efficiency relationship to account for this fact:
\begin{equation}
\eta_{weak}(M|N,P|Q,S) = \frac{T(M,Q)}{S \times T(N,P)}\:,
  \label{eq:ms_weak_efficiency}
\end{equation}
where now the absolute efficiency is a function of the number of sets
in the problem and again $M/Q = N/P$. As with the strong scaling
study, we will use the 16-core single set case as the reference
computation for efficiencies. Interestingly,
Eq~(\ref{eq:ms_weak_efficiency}) has a very similar form to the strong
scaling efficiency given by Eq~(\ref{eq:strong_scaling_absolute}),
representative of the strong scaling nature of adding additional
sets. For problems where adding sets improves time to solution, this
new weak scaling formulation accounts for how efficiently those extra
resources are incorporated into the problem.

@@

It was found the MCSA can indeed be effectively parallelized using the
multiple-set overlapping-domain decomposition algorithm borrowed from
the reactor physics community.  The new algorithm was tested in a wide
variety of parallel scaling studies on the Titan Cray XK7 machine at
the Oak Ridge Leadership Computing Facility. To test the algorithm at
high levels of concurrency, up to 65,356 cores were used in strong
scaling exercises and 131,072 cores used in weak scaling exercises
using a neutron diffusion problem.

@@

For future work, overall parallel scaling results indicated that the
implementation was of good quality in both the strong and weak scaling
cases. However, for the multiple set cases in the weak scaling
analysis, it was noted that there was a growing parallel overhead as a
function of core count. This was counter-intuitive in that the size of
the parallel group over which the set reductions were occurring was
fixed at all core counts and therefore we do not expect the addition
of such an operation to increase parallel run times. It did, however,
and therefore future research should explore why the implementation
for multiple sets generated parallel overhead that increased with core
count instead of remaining constant.
